\thispagestyle{empty}
{\bf \asharpName{} User Guide}


Copyright \copyright 1990-2007 Aldor Software Organization Ltd (Aldor.org).

\begin{small}
All rights reserved.  No part of this Manual may be reproduced, 
transcribed, stored in a retrieval system, translated into any language 
or computer language or transmitted in any form or by any means, 
electronic, mechanical, photocopying, recording or otherwise, without 
the prior written permission of the copyright owner.

The copyright owner gives no warranties and makes no representations 
about the contents of this Manual and specifically disclaims any implied
warranties of merchantability or fitness for any purpose.

The copyright owner reserves the right to revise this Manual and to make
changes from time to time in its contents without notifying any person 
of such revisions or changes.

\end{small}

%*****************************************************************

Substantial portions of this manual have been published earlier
in the volume ``Axiom Library Compiler User Guide,'' 
The Numerical Algorithms Group 1994, {\tt ISBN 1-85206-106-5}.

\begin{small}
\asharpName{} was originally developed, under the working name
of $A^\sharp$, by the Research Division of
International Business Machines Corporation,
Yorktown Heights, New York, USA.

Aldor, AXIOM and the AXIOM logo are trademarks of NAG.
\\
NAG is a registered trademark of the Numerical Algorithms Group Limited.
%\vspace{-\baselineskip}
All other trademarks are acknowledged.
\end{small}
\vspace{\baselineskip}

\clearpage

